---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 7 - Due date 03/20/23"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A07_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Set up

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(lubridate)
library(ggplot2)
library(tidyverse)
library(Kendall)
```

## Importing and processing the data set

Consider the data from the file "Net_generation_United_States_all_sectors_monthly.csv". The data corresponds to the monthly net generation from January 2001 to December 2020 by source and is provided by the US Energy Information and Administration. **You will work with the natural gas column only**.

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

### Q1

Import the csv file and create a time series object for natural gas. Make you sure you specify the **start=** and **frequency=** arguments. Plot the time series over time, ACF and PACF.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r}
#import data
ng_raw <- read.csv("/Users/karenthornton/Documents/School/Grad School/Year 2/Semester 2/Time Series/TimeSeriesAnalysis_Sp23/Data/Net_generation_United_States_all_sectors_monthly.csv", skip=4)

ng_raw$Month <- mdy(ng_raw$Month)

ng_selected <- select(ng_raw, Month, natural.gas.thousand.megawatthours)

colnames(ng_selected) <- c("Month", "Natural_Gas")
#create timeseries 
ng_ts <- ts(ng_selected$Natural_Gas, frequency = 12, start = 1)
```

```{r}
#plot
ggplot(ng_selected)+
  geom_line(aes(x=Month, y=Natural_Gas))

#ACF
Acf(ng_ts,lag.max = 40,main="Natural Gas ACF",ylim=c(-1,1))

#PACF
Pacf(ng_ts,lag.max = 40,main="Natural Gas PACF",ylim=c(-1,1))
```

### Q2

Using the $decompose()$ or $stl()$ and the $seasadj()$ functions create a series without the seasonal component, i.e., a deseasonalized natural gas series. Plot the deseasonalized series over time and corresponding ACF and PACF. Compare with the plots obtained in Q1.

```{r}
ng_decompose <- decompose(ng_ts, type="additive")
plot(ng_decompose)

ng_deaseasonal_decompose <- seasadj(ng_decompose)
```


```{r}
#plot
ggplot(ng_selected, aes(x=Month))+
  geom_line(aes(y=ng_deaseasonal_decompose))

#ACF
Acf(ng_deaseasonal_decompose,lag.max = 40,main="Deseasonal Natural Gas ACF",ylim=c(-1,1))

#PACF
Pacf(ng_deaseasonal_decompose,lag.max = 40,main="Deseasonal Natural Gas PACF",ylim=c(-1,1))
```
#The plots here show less seasonality than the ones in question 1. The initial graph shows an increasing trend, but doesn't have the same ups and downs as the initial plot in quesiton 1. The ACF and PACF here show no seasonality, which is different than the seasonality that is shown in the ACF and PACF for quesiton 1. 

## Modeling the seasonally adjusted or deseasonalized series

### Q3

Run the ADF test and Mann Kendall test on the deseasonalized data from Q2. Report and explain the results.

```{r}
print(adf.test(ng_deaseasonal_decompose))
print(MannKendall(ng_deaseasonal_decompose))
```
#The p-value for both the adf test and the MannKendall are both significant (p value is less than 0.05).The ADF test show that there isn't a unit root and that it is stationary. The MannKendall test shows that there's a trend.

### Q4

Using the plots from Q2 and test results from Q3 identify the ARIMA model parameters $p,d$ and $q$. Note that in this case because you removed the seasonal component prior to identifying the model you don't need to worry about seasonal component. Clearly state your criteria and any additional function in R you might use. DO NOT use the $auto.arima()$ function. You will be evaluated on ability to can read the plots and interpret the test results.

```{r}
#need to difference because even after deseasonalizing it there's still a slow decay in the ACF. 
n_diff <- ndiffs(ng_deaseasonal_decompose)

ng_diff <- diff(ng_deaseasonal_decompose, differences = 1)

#ACF
Acf(ng_diff,lag.max = 40,main="Differenced Natural Gas ACF",ylim=c(-1,1))

#PACF
Pacf(ng_diff,lag.max = 40,main="Differenced Natural Gas PACF",ylim=c(-1,1))

```
#Answer: (1,1,1) Based on the de-seasonal time series there's a slow decay in the ACF. This means that it's most likely an AR model. It is harder to tell what it is when it is differenced. The d is equal to 1 because you have to difference it once. You know you have to difference it at least once because there isn't any unit root. The p value is 1 because looking at the PACF there's only one significant lag after lag zero. The q value is 1 because looking at the ACF there's only one significant lag after lag zero. 

### Q5

Use $Arima()$ from package "forecast" to fit an ARIMA model to your series considering the order estimated in Q4. You should allow constants in the model, i.e., $include.mean = TRUE$ or $include.drift=TRUE$. **Print the coefficients** in your report. Hint: use the $cat()$ function to print.

```{r}
Model_111 <- Arima(ng_diff,order=c(1,1,1),include.drift=TRUE)
print(Model_111)
```

### Q6

Now plot the residuals of the ARIMA fit from Q5 along with residuals ACF and PACF on the same window. You may use the $checkresiduals()$ function to automatically generate the three plots. Do the residual series look like a white noise series? Why?

```{r}
checkresiduals(Model_111)
```

#Yes, the series looks like white noise. White noise means the series is random and you can't predict it. Looking at the residual graph you can see that the points are random and there's no way to predict it. You can also see that the ACF values are all close to zero. This means that the points aren't correlated to each other (other than the random coorelation). 

## Modeling the original series (with seasonality)

### Q7

Repeat Q4-Q6 for the original series (the complete series that has the seasonal component). Note that when you model the seasonal series, you need to specify the seasonal part of the ARIMA model as well, i.e., $P$, $D$ and $Q$.

```{r}
print(adf.test(ng_ts))

n_diff_2 <- ndiffs(ng_ts)

ng_diff_2 <- diff(ng_ts, differences = 1)

#ACF
Acf(ng_diff_2,lag.max = 40,main="Original Natural Gas ACF",ylim=c(-1,1))

#PACF
Pacf(ng_diff_2,lag.max = 40,main="Original Natural Gas PACF",ylim=c(-1,1))
```
#Answer: (1,1,1)(2,1,2)[12] The d for this is 1 because the adf test shows you need at least one difference since there isn't a unit root. The ndiff function shows how many times you have to difference it.This is the same as question 4 because that part is looking at the non-seasonal part. The D for this is 1 because the adf test for the non-seasonal series also has a significant p-value, meaning there isn't a root unit and it needs to be differenced at least once. the ndiff function shows only 1 is needed.  Looking at the ACF you can see that q is equal to 1 because that's the first significant lag after lag 0. Looking at the PACF you can see that p is equal to 1 as well because lag 1 is the first significant lag after lag zero. You can tell that Q is 2 because looking at the ACF, there are two siginificant lags after the first seasonal spike (around lag 11). P is also 2 becuase if you look at the PACF there are two significant negative spikes after the first lag. 

```{r}
Model_seasonal <- Arima(ng_diff,order=c(1,1,1), seasonal = c(2,1,2),include.drift=TRUE)
print(Model_seasonal)
```

```{r}
checkresiduals(Model_seasonal)
```
#The residuals here look slightly less like white noise. If you look at the ACF there are some ups and downs even though the coorelations are close to 0. This is because the seasonality is affecting the coorelation and thus is not a white noise.


### Q8

Compare the residual series for Q7 and Q6. Can you tell which ARIMA model is better representing the Natural Gas Series? Is that a fair comparison? Explain your response.

```{r}
checkresiduals(Model_111)
checkresiduals(Model_seasonal)
```
#The deseasonaed ARIMA model is better representing the Natural Gas Series because it is closer to white noise. This means that the model that we fit is actually working. The ACF for the residuals with the seasonality still in there is making the series not equal to white noise. This is a fair comparision because you need to take the seasonality (if there is one) out of a series in order to see the true coorelation. Taking the seasonality out of the time series make the model fit better and thus looks closer to white noise.


## Checking your model with the auto.arima()

**Please** do not change your answers for Q4 and Q7 after you ran the $auto.arima()$. It is **ok** if you didn't get all orders correctly. You will not loose points for not having the same order as the $auto.arima()$.

### Q9

Use the $auto.arima()$ command on the **deseasonalized series** to let R choose the model parameter for you. What's the order of the best ARIMA model? Does it match what you specified in Q4?

```{r}
auto.arima(ng_deaseasonal_decompose)
```
#This does not match what I got. In this, p is 3, d is 1, and q is 0. This means that I got d correct, but not p or q. 

### Q10

Use the $auto.arima()$ command on the **original series** to let R choose the model parameters for you. Does it match what you specified in Q7?

```{r}
auto.arima(ng_ts)
```
#I got the seasonal part correct (2,1,2) but not the non-seasonal part. For this I got (1,1,1) and it should be (2,0,1). I'm not sure why d is 0 since the adf test was significant. s is 12 because it's monthly data.
