---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 6 - Due date 03/06/23"
author: "Karen Thornton"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

## Questions 

This assignment has general questions about ARIMA Models. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
#install.packages("sarima")
library(sarima)
library(forecast)
library(tseries)
library(ggplot2)
library(Kendall)
```


## Q1
Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models: 


* AR(2)

> Answer: With an autoregressive model with order 2, it is important to look at the PACF. The ACF for AR models decay exponentially with time. The PACF will show that the order is 2. In this case lag 2 will be the last lag that is significant (above the significance line). 

* MA(1)

> Answer: With a moving average, order 1, it is important to look at the ACF. The PACF for moving averages will decay exponentially. The ACF will show the order. For this case lag 1 will be the cutoff. This means that lag one will be the last significant lag (last above the significance line). Usually for MAs the cut off has a negative autocorrelation. 



## Q2
Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$. Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use R to generate $n=100$ observations from each of these three models

```{r}
#ARMA(1,0)

ARMAmodel_1<- arima.sim(model=list(ar=0.6), n=100) #the AR coefficient is 0.6

#ARMA(0,1)

ARMAmodel_2<- arima.sim(model=list(ma=0.9), n=100) #the MA coefficient is 0.9

#ARMA(1,1)

ARMAmodel_3<- arima.sim(model=list(ma=0.9, ar=0.6), n=100)
```


(a) Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r}
par(mfrow=c(1,3))
  Acf(ARMAmodel_1,lag.max = 40,main="ARMA(1,0)",ylim=c(-1,1))
  Acf(ARMAmodel_2,lag.max = 40,main="ARMA(0,1)",ylim=c(-1,1))
  Acf(ARMAmodel_3,lag.max = 40,main="ARMA(1,1)",ylim=c(-1,1))
```


(b) Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r}
par(mfrow=c(1,3))
  Pacf(ARMAmodel_1,lag.max = 40,main="ARMA(1,0)",ylim=c(-1,1))
  Pacf(ARMAmodel_2,lag.max = 40,main="ARMA(0,1)",ylim=c(-1,1))
  Pacf(ARMAmodel_3,lag.max = 40,main="ARMA(1,1)",ylim=c(-1,1))
```

(c) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer:It would be hard to identify these because of the seasonality, but here's what you can tell:The ACFs for ARMA (1,0) and ARMA (1,1) show slightly more exponential decay than the ARMA (0,1) which makes sense since AR models have more exponential decay on their ACFs than MA models. If you look at the PACFs the ARMA(1,0) has a lag that's close to one which is close to the given lag. The PACF for the ARMA (0,1) slightly decays exponentially and the ACF shows a lag around 1 as well. The ARMA (1,1) is harder to identify, but the ACF has a strong exponential decay and the PACF has a slight decay you can tell that it has AR and MA. It would be hard to identify the order for the ARMA (1,1)

(d) Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: The coefficient for AR is around 0.5 (you can tell from the lag on the PACF) which is close to the given 0.6. The coefficient for the MA is also around 0.5 (you can tell from the ACF) which isn't that close to 0.9. This may be because I set the number of lags to 40 which might limit the number of lags that would optimize the model. This cold also be due to the limit in number of observations. The MA could be off because the lag is just not shown on the graph. You can tell that the order is 1 but you can't tell what the coefficient is. 


(e) Increase number of observations to $n=1000$ and repeat parts (a)-(d).

```{r}
#ARMA(1,0)

ARMAmodel_4<- arima.sim(model=list(ar=0.6), n=1000) #the AR coefficient is 0.6

#ARMA(0,1)

ARMAmodel_5<- arima.sim(model=list(ma=0.9), n=1000) #the MA coefficient is 0.9

#ARMA(1,1)

ARMAmodel_6<- arima.sim(model=list(ma=0.9, ar=0.6), n=1000)

par(mfrow=c(1,3))
  acf(ARMAmodel_4,lag.max = 40,main="ARMA(1,0)",ylim=c(-1,1))
  acf(ARMAmodel_5,lag.max = 40,main="ARMA(0,1)",ylim=c(-1,1))
  acf(ARMAmodel_6,lag.max = 40,main="ARMA(1,1)",ylim=c(-1,1))
  
par(mfrow=c(1,3))
  pacf(ARMAmodel_4,lag.max = 40,main="ARMA(1,0)",ylim=c(-1,1))
  pacf(ARMAmodel_5,lag.max = 40,main="ARMA(0,1)",ylim=c(-1,1))
  pacf(ARMAmodel_6,lag.max = 40,main="ARMA(1,1)",ylim=c(-1,1))
```

>Answer: In this case it is easier to tell that the ARMA (1,0)
 and ARMA (1,1) have an exponential decay in the ACF. The ARMA (0,1) and
 ARMA (1,1) also have expoenential decay in the PACF. 
 
>Answer: The coefficient for ARMA (1,0) is around 0.6 and the
coefficient for ARMA (0,1) is around 0.9. These are the same as the given coefficients. They might be more correct than the previous ones because there's more observations. 
 
## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$


(a) Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

The integers are ARIMA$(1,1,1)(1,1,0)_12$

So 
p=1, d=1, q=1, P=1, D=1, Q=0, s=12

p is 1 because its Yt-1 (not Yt-2 or beyond)
d is 1 because there's no constant term which means the series is differenced 
q is 1 because its At-1 (not At-2 or beyond)
s is 12 because there's a Yt-12 component to show seasonality (12 months in a year)
P=1 because there's Yt-12 (12 being the s)
Q=0 because there isn't At-12 (no seasonality)
D=1 because there isn't a constant

#no constant term, so the series has been differenced, so d=1 and D=1

(b) Also from the equation what are the values of the parameters, i.e., model coefficients. 

The AR coefficient is 0.7
The MA coefficient is 0.1
The seasonal AR coefficient is -0.25

## Q4

Plot the ACF and PACF of a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using R. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot ACF and PACF for the simulated data. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r}
SARIMAmodel<- sim_sarima(model=list(ma=0.5,sar=0.8, nseasons=12), n=1000) 

par(mfrow=c(1,2))
  acf(SARIMAmodel,lag.max = 40,main="SARIMA_acf",ylim=c(-1,1))
  pacf(SARIMAmodel,lag.max = 40,main="SARIMA_pacf",ylim=c(-1,1))
```
>Answer: You can tell from the ACF that the s is 12 becuase you can see the seasonal spike around that time. Looking at the non-seasonal lags the ACF is slowly decaying which means AR process. The PACF non-seasonal lag cuts off which also means AR. The q value is 1 and the p value is harder to tell but I thik its also 1. Looking at seasonality, the ACF has multiple spikes at each seasonality so it's an AR and the PACF only has one spike at the seasonal lags wich confirms AR. The P and Q are hard to tell but I think P=1 and Q=1 The graphs show what was just written, but the equation shows that p=0, q=1, because there was no given AR coefficient. The equation also shows that ther sould be P=1 and Q=0 because the sar coefficent is identified but there was no given sma coefficient. Because of all this i think the graphs are hard to show what the equation shows. 
