---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 3 - Due date 02/10/23"
author: "Karen Thornton"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

```{r}
#Data set
Renewable_Energy <- read.csv(file = "/Users/karenthornton/Documents/School/Grad School/Year 2/Semester 2/Time Series/TimeSeriesAnalysis_Sp23/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv", header = TRUE, dec = ".", sep=",",stringsAsFactors = TRUE)

library(dplyr)
Renewable_Energy <- select(Renewable_Energy, c("Month","Total.Biomass.Energy.Production",
                        "Total.Renewable.Energy.Production",
                        "Hydroelectric.Power.Consumption"))


library(lubridate)
Renewable_Energy$Month <- ym(Renewable_Energy$Month)
nenergy <- ncol(Renewable_Energy)-1
nobs <- nrow(Renewable_Energy) 
ts_Renewable_Energy <- ts(Renewable_Energy[,2:(2+nenergy-1)],
                           frequency = 12, start = c(1973,1))
```

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)

```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)

```{r}
library(ggplot2)
name <- colnames(Renewable_Energy)

par(mfrow=c(1,3))
for(i in 1:nenergy){
  acf(ts_Renewable_Energy[,i],lag.max = 40,main=name[i+1],ylim=c(-1,1))
  pacf(ts_Renewable_Energy[,i],lag.max = 40,main=name[i+1],ylim=c(-1,1))
  plot(ts_Renewable_Energy[,i])+
    title(main = paste(name[1+i]))
}
```

### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

#Answer:
Yes, all three of them seem to have a trend. Biomass and Renewable Energy have a positive linear trend and Hydroelectric has a negative linear trend (this trend is less strong than the other two).It also looks like the Hydroelectric Power consumption has a seasonal trend.

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r}
#Create vector t
t <- c(1:nobs)

linear_trend_1 <- lm(Renewable_Energy[,2]~t)
linear_trend_2 <- lm(Renewable_Energy[,3]~t)
linear_trend_3 <- lm(Renewable_Energy[,4]~t)

#for (i in 1:nenergy) {
 # linear_trend <- lm(Renewable_Energy[,i+1] ~ t)
#}
summary(linear_trend_1)
summary(linear_trend_2)
summary(linear_trend_3)

#interpret the regression output: The intercept is the first coefficient and the slope is the second coefficient

beta0_1 <- as.numeric(linear_trend_1$coefficients[1])
beta1_1 <- as.numeric(linear_trend_1$coefficients[2])
beta0_2 <- as.numeric(linear_trend_2$coefficients[1])
beta1_2 <- as.numeric(linear_trend_2$coefficients[2])
beta0_3 <- as.numeric(linear_trend_3$coefficients[1])
beta1_3 <- as.numeric(linear_trend_3$coefficients[2])
```

```{r}
#plot for visual aid
for (i in 1:nenergy) {
  print(ggplot(Renewable_Energy, aes(x=Month, y=Renewable_Energy[,(i+1)])) +
          geom_line(color="blue") +
          ylab(paste0("Inflow ",colnames(Renewable_Energy)[(i+1)],sep="")) +
          geom_smooth(color="red",method="lm"))
}
```
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r}

  detrend_Renewable_Energy_1 <- Renewable_Energy[,2]-(beta0_1+beta1_1*t)

  print(ggplot(Renewable_Energy, aes(x=Month, y=Renewable_Energy[,2]))+
    geom_line(color="blue")+
    geom_line(aes(y=detrend_Renewable_Energy_1), color="green")+
      ylab(paste0("Energy ",colnames(Renewable_Energy)[2],sep="")))
  
   detrend_Renewable_Energy_2 <- Renewable_Energy[,3]-(beta0_2+beta1_2*t)

  print(ggplot(Renewable_Energy, aes(x=Month, y=Renewable_Energy[,3]))+
    geom_line(color="blue")+
    geom_line(aes(y=detrend_Renewable_Energy_2), color="green")+
      ylab(paste0("Energy ",colnames(Renewable_Energy)[3],sep="")))
  
   detrend_Renewable_Energy_3 <- Renewable_Energy[,4]-(beta0_3+beta1_3*t)

  print(ggplot(Renewable_Energy, aes(x=Month, y=Renewable_Energy[,4]))+
    geom_line(color="blue")+
    geom_line(aes(y=detrend_Renewable_Energy_3), color="green")+
      ylab(paste0("Energy ",colnames(Renewable_Energy)[4],sep="")))

```

#Answer:
The detrended lines look different than the original, especially for biomass and renewable energy (hydroelectric stayed mostly the same shape which means the trend was weak). The mean is closer to zero for the detrended lines. 

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?

```{r}
  par(mfrow=c(1,2))
  acf(detrend_Renewable_Energy_1,lag.max = 40,main=name[2],ylim=c(-1,1))
  pacf(detrend_Renewable_Energy_1,lag.max = 40,
       main=paste0("Biomass"),ylim=c(-1,1))
  
  par(mfrow=c(1,2))
  acf(detrend_Renewable_Energy_2,lag.max = 40,main=name[3],ylim=c(-1,1))
  pacf(detrend_Renewable_Energy_2,lag.max = 40,
       main=paste0("Renewable Energy"),ylim=c(-1,1))
  
  par(mfrow=c(1,2))
  acf(detrend_Renewable_Energy_3,lag.max = 40,main=name[4],ylim=c(-1,1))
  pacf(detrend_Renewable_Energy_3,lag.max = 40,
       main=paste0("Hydroelectric"),ylim=c(-1,1))
```

#Answer:
It doesn't look like the plots have changed too much. The only thing I notice is the Hydroelectric ACF looks even more seasonal than before. 

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

```{r}
#The Hydroelectric seems to have a seasonal trend

#First create the seasonal dummies
dummies <- seasonaldummy(ts_Renewable_Energy[,3])  

#Then fit a linear model to the seasonal dummies
seas_means_model=lm(Renewable_Energy[,(4)]~dummies)
summary(seas_means_model)

#Store regression coefficients
beta_int=seas_means_model$coefficients[1]
beta_coeff=seas_means_model$coefficients[2:12]

#compute seasonal component
Energy_seas_comp=array(0,nobs)
for(i in 1:nobs){
  Energy_seas_comp[i]=(beta_int+beta_coeff%*%dummies[i,])
}
```

```{r}
#plot
ggplot(Renewable_Energy, aes(x=Month, y=Renewable_Energy[,4])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(Renewable_Energy)[(4)],sep="")) +
            geom_line(aes(y=Energy_seas_comp), col="red")
```
### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

```{r}
#Removing seasonal component
deseason_Renewable_Energy <- Renewable_Energy[,4]-Energy_seas_comp

#plot
ggplot(Renewable_Energy, aes(x=Month, y=Renewable_Energy[,4])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(Renewable_Energy)[4],sep="")) +
            geom_line(aes(y=deseason_Renewable_Energy), col="green")
```

### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?

```{r}
  par(mfrow=c(1,2))
  acf(deseason_Renewable_Energy,lag.max = 40,main=name[4],ylim=c(-1,1))
  pacf(deseason_Renewable_Energy,lag.max = 40,
       main=paste0("Hydroelectric"),ylim=c(-1,1))
```

#Answer:
Yes, the plots from Q8 are different than Q1. Since taking the seasonal trend out, the hydroelectric power consumption ACF doesn't look seasonal (no ups and downs). 